---
title: "BCB420 Assignment #1: Data set selection, cleaning, and normalization"
author: "Yuxi Zhang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    code_folding: hide
    self_contained: true
    number_sections: true
bibliography: references.bib
link-citations: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,   # Hides "Loading required package" and "Attaching package"
  warning = FALSE,   # Hides warning messages
  comment = "" # Keeps your table results visible
)
```

# Dataset Selection: GSE234357
The initial dataset for this project was GSE255720, a bulk RNA-seq study profiling colon biopsies from individuals with Inflammatory Bowel Disease (IBD) and non-IBD controls. This dataset included sufficient biological replication (five samples per condition) and was approved by Prof. Isserlin from the dataset collection homework [@arifuzzaman2024]. 

However, upon inspection of the GEO submission, I found that gene-level raw count matrices were not deposited for GSE255720. The authors provided only a pre-normalized expression matrix at the GSE level. Supplemental GSM-level directories were empty, and raw read count files suitable for standard RNA-seq normalization and differential expression workflows. I then chose to conduct analysis for a scRNA-seq dataset from the **same publication**, but the flaw with that dataset was it contained a single replicate [@arifuzzaman2024]. 

Finally, I selected the **GSE234357** dataset [@jarosch2023]. 

The GSE234357 dataset investigates the complex pathophysiology of acute graft-versus-host disease (aGvHD) in individuals following allogeneic hematopoietic stem cell transplantation (aHSCT). By employing a multimodal approach—integrating single-cell RNA sequencing (scRNA-seq) with DNA-barcoded antibodies and in situ multiplexed tissue imaging (ChipCytometry)—the researchers analyzed gastrointestinal (GI) biopsies to map the immune landscape of the gut. The study specifically sought to define the spatiotemporal distribution of T cell infiltrates and integrate this with 16S microbiome data and clinical biomarkers. Key findings from this cohort highlight that high microbiome diversity and short-chain fatty acid-producing bacteria are linked to the sustenance of suppressive regulatory T cells (Tregs), while severe aGvHD is driven by the clonal expansion of CD8 T cells [@jarosch2023, @davis2007].

* **Publication Link:** [Multimodal immune cell phenotyping in GI biopsies reveals microbiome-related T cell modulations in human GvHD](https://doi.org/10.1016/j.xcrm.2023.101125)
* **Conditions:** Colon biopsies from aGvHD patients vs. Colon biopsies from acute GvHD patients 

---
# Cleaning and Analyzation Procedure 

## Clean the data and map to HUGO symbols 

### Set up packages
```{r, results='hide'}
# Setup block: Check and install missing packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

packages <- c("GEOquery", "edgeR", "org.Hs.eg.db", "pheatmap", "ggrepel", "tidyverse")

for (p in packages) {
    if (!requireNamespace(p, quietly = TRUE)) {
        if (p %in% rownames(available.packages())) {
            install.packages(p)
        } else {
            BiocManager::install(p)
        }
    }
}
library(GEOquery)
library(Biobase)
library(Matrix)
library(tidyverse)
library(org.Hs.eg.db)
library(pheatmap)
library(DESeq2)
library(edgeR)
library(ggplot2)
library(ggrepel) 

```
The pipeline setup utilizes BiocManager to acquire specialized Bioconductor packages [@huber2015] such as org.Hs.eg.db [@carlson2024] for genome annotation and edgeR [@edger2025] or DESeq2 for differential expression analysis. General data science tools from the tidyverse [@wickham2019] are also integrated to facilitate efficient data manipulation and visualization through ggplot2 [@slowikowski2024] and pheatmap [@kolde2025].

### Download the data with the GEOquery Bioconductor package


```{r, results='hide'}
library(GEOquery)

gse_id <- "GSE234357"
base_dir <- "GSE234357"
raw_tar_path <- file.path(base_dir, gse_id, paste0(gse_id, "_RAW.tar"))
extract_dir <- file.path(base_dir, "extracted")

# 1. Only extract raw data if the tar exists locally
if (file.exists(raw_tar_path)) {
  message("Raw files exist locally, attempting extraction...")
  if (!dir.exists(extract_dir)) dir.create(extract_dir, recursive = TRUE)
  tryCatch(
    untar(raw_tar_path, exdir = extract_dir),
    error = function(e) message("Skipping extraction: ", e$message)
  )
} else {
  message("Raw files not found. Skipping extraction (Docker-friendly).")
}

# 2. GEO metadata
geo_dir <- "geo_data"
if (!dir.exists(geo_dir)) dir.create(geo_dir, recursive = TRUE)

gse <- getGEO(gse_id, destdir = geo_dir, GSEMatrix = TRUE)
metadata <- pData(gse[[1]])

# 3. Filter metadata
metadata_expr <- metadata[grep("gene expression", metadata$title), ]
metadata_expr$Condition <- ifelse(grepl("GvHD", metadata_expr$title), "GvHD", "Control")
```
Data acquisition is performed using the GEOquery package to programmatically retrieve the GSE234357 dataset from the Gene Expression Omnibus [@davis2007]  .

### Prepare Dataset {#Prepare-Dataset}

```{r}
# Initialize the list to store counts
pseudobulk_list <- list()

# Loop only through the 9 Gene Expression samples (metadata_expr from previous step)
for (i in 1:nrow(metadata_expr)) {
  gsm_id <- metadata_expr$geo_accession[i]
  
  # 1. Identify files for this specific GSM
  # We look for files starting with the GSM ID and ending with the 10x suffixes
  m_file <- list.files("GSE234357/extracted", pattern = paste0(gsm_id, ".*matrix.mtx.gz"), full.names = TRUE)
  f_file <- list.files("GSE234357/extracted", pattern = paste0(gsm_id, ".*features.tsv.gz"), full.names = TRUE)
  
  if (length(m_file) == 1 && length(f_file) == 1) {
    message("Processing ", gsm_id, "...")
    
    # 2. Read the Sparse Matrix and Features
    mat <- readMM(m_file)
    feat <- read.table(f_file, header = FALSE, sep = "\t")
    
    # 3. Aggregate: Sum all cells (columns) to get total counts per gene (rows)
    # This creates the 'pseudobulk' profile for that sample pool
    pool_counts <- rowSums(mat)
    
    # 4. Use Ensembl IDs (V1) as names for alignment
    names(pool_counts) <- feat$V1
    
    # Add to pseudobulk list
    pseudobulk_list[[gsm_id]] <- pool_counts
  } else {
    warning("Could not find unique files for ", gsm_id)
  }
}

# 5. Merge into a single Raw Count Dataframe
raw_counts <- as.data.frame(do.call(cbind, pseudobulk_list))

# Preview the result
knitr::kable(head(raw_counts[, 1:5]), caption = "Table 1: Data Pre-processing and Annotation Summary. Summary of the bioinformatics pipeline's data cleaning steps. The table tracks the transition from raw Ensembl IDs to unique HUGO gene symbols, including the percentage of data filtered as non-coding or unmapped noise (33.17%) and the resolution of many-to-one mappings.")
```
The raw data is processed into a "pseudobulk" format to transition from single-cell resolutions to sample-level counts [@pseudobulk2020]. 
* For each GSM identifier, the method reads the sparse count matrix and corresponding feature lists, then calculates the rowSums across all individual cells within a sample pool. 
* These summarized counts are then merged into a single comprehensive dataframe using Ensembl IDs as the primary indices, providing a high-level overview of the raw transcriptomic landscape across all experimental replicates. 

```{r, results='hide'}
# Check total library sizes per pool
colSums(raw_counts)
```


### Map to HUGO symbols {#Map-to-HUGO-symbols}


```{r, results='hide'}
# 1. Define Metadata
metadata_clean <- metadata %>%
  dplyr::select(geo_accession, title) %>%
  filter(grepl("gene expression", title)) %>%
  mutate(Group = case_when(
    geo_accession %in% c("GSM7465905", "GSM7465907", "GSM7465911") ~ "Control",
    TRUE ~ "GvHD"
  ),
  Batch = gsub("_Pool.*", "", title))

# 2. Unified Mapping
mapping <- AnnotationDbi::select(org.Hs.eg.db,
                                 keys = rownames(raw_counts),
                                 keytype = "ENSEMBL",
                                 columns = "SYMBOL")

mapping_unique <- as.data.frame(mapping) %>%
  filter(!is.na(SYMBOL)) %>% 
  group_by(ENSEMBL) %>%
  dplyr::slice(1) %>% 
  ungroup()

# 3. Create 'clean_counts' 
clean_counts <- raw_counts %>%
  tibble::rownames_to_column("ENSEMBL") %>%
  dplyr::inner_join(mapping_unique, by = "ENSEMBL") %>%
  dplyr::select(-ENSEMBL) %>%
  dplyr::group_by(SYMBOL) %>%
  dplyr::summarise(across(everything(), max)) %>%
  tibble::column_to_rownames("SYMBOL")

# 4. Create 'qc_summary'
qc_summary <- data.frame(
  Sample = colnames(clean_counts),
  Total_Reads = colSums(clean_counts)
) %>% merge(metadata_clean, by.x = "Sample", by.y = "geo_accession")
```
To ensure biological interpretability, Ensembl identifiers are mapped to standardized HUGO gene symbols using the org.Hs.eg.db SQLite database [@carlson2024]. 
* The cleaning process involves filtering out unmapped "noise" (non-coding or outdated IDs) and resolving many-to-one mapping redundancies [@matrix2025]. 
* By employing a "Max-Expression" strategy—where the highest expression value is retained for redundant symbols—the data is collapsed into a final, unique gene-by-sample matrix. 
* This step concludes with the generation of a quality control (QC) summary that links total read counts to the cleaned clinical metadata.

### Pre-normalization Boxplot {#Pre-normalization-Boxplot} 
```{r, fig.cap="Figure 1: Pre-normalization Gene Expression Distribution. Boxplots representing the distribution of $Log_2(Counts + 1)$ across nine gastrointestinal biopsy samples. The vertical labels represent Sample IDs (GSM), while colors denote biological groups (Skyblue: Control, Salmon: GvHD). The red dashed line indicates the global median. Significant technical variance and baseline shifts are visible, necessitating normalization."}
# 1. Prepare the numeric data
numeric_counts <- as.matrix(clean_counts)
log_counts <- log2(numeric_counts + 1)

# 2. Set margins: bottom, left, top, right (8 lines for the bottom)
par(mar=c(8, 4, 4, 2)) 

# 3. Define colors based on Group
# Ensure qc_summary order matches clean_counts columns
box_cols <- ifelse(qc_summary$Group == "Control", "skyblue", "salmon")

# 4. Plot
boxplot(log_counts,
        main = "Pre-normalization Distribution",
        ylab = "Log2(Counts + 1)",
        col = box_cols, 
        las = 2,        # Rotates axis labels 90 degrees
        cex.axis = 0.7) # Shrinks font slightly to fit labels

# 5. Add the 'Samples' label lower down so it doesn't overlap
title(xlab="Samples", line=6)

# 6. Add global median line
abline(h = median(log_counts), col = "red", lwd = 2, lty = 2)
```
In the initial pre-normalization plot, gene expression is represented as $Log_2(Counts + 1)$. There is a clear lack of alignment among the samples, as evidenced by the varying positions of the black median bars relative to the red dashed global median line. Samples like GSM7465917 show a much broader distribution and a higher median compared to samples like GSM7465903 or GSM7465909. These discrepancies reflect significant technical noise and differences in sequencing depth—specifically the 25-fold difference in library sizes observed in the raw counts—which must be corrected to prevent biased results in downstream differential expression analysis.

## Apply Normalization 

The normalization step is essential to account for differences in library sizes and RNA composition bias between samples. After converting the cleaned counts into a numeric matrix and initializing a DGEList object, I applied the Trimmed Mean of M-values (TMM) scaling method via the calcNormFactors function. This approach ensures that the expression levels are comparable across the nine patient pools by calculating scaling factors that adjust the effective library sizes. The success of this process is visualized using a boxplot of $log_2$ counts per million (CPM), which demonstrates that the global medians of the samples are aligned, effectively removing technical baseline shifts [@robinson2010].

```{r, fig.cap="Figure 2: Post-normalization Distribution (TMM). Boxplots showing the distribution of $Log_2$ CPM after Trimmed Mean of M-values (TMM) normalization. The blue dashed line represents the global median. The alignment of the boxes demonstrates the successful removal of library size and composition bias, allowing for direct comparison of biological variance."}
# 1. Ensure the DGEList is built from numeric data only
# This prevents the 'non-numeric' error downstream
numeric_counts <- as.matrix(clean_counts)
dge <- DGEList(counts = numeric_counts, group = factor(qc_summary$Group))

# 2. Apply TMM Normalization
dge <- calcNormFactors(dge)

# 3. Calculate Normalized Log2 CPM
norm_counts <- cpm(dge, log = TRUE)


# 4. Figure 2: Post-normalization Distribution (TMM)
par(mar=c(8, 4, 4, 2)) # Match the 8-line bottom margin
boxplot(norm_counts,
        main = "Post-normalization Distribution (TMM)",
        ylab = "Log2 CPM",
        col = box_cols, # Using the Skyblue/Salmon colors defined earlier
        las = 2, 
        cex.axis = 0.7)

# Move the X-axis label down to avoid overlapping the GSM IDs
title(xlab="Samples", line=6)

# Add a blue dashed line at the global median to show alignment
abline(h = median(norm_counts), col = "blue", lwd = 2, lty = 2)
```

Following TMM normalization, the data is represented as $Log_2$ CPM (Counts Per Million).The median values for all nine samples now sit more closely on the blue dashed global median line. This indicates that the normalization process has effectively leveled the playing field by calculating scaling factors to account for library size and composition bias. For instance, samples that were initially under-sequenced were scaled up, while those dominated by highly expressed genes were scaled down to ensure mathematical comparability [@robinson2010].

Despite the alignment of the medians, the plots retain a distinct "bottom-heavy" asymmetry, where the bulk of the data (the colored boxes) is situated at the lower end of the y-axis with numerous high-expression outliers extending upward. This is a common and expected characteristic of transcriptomic data, reflecting the biological reality that the majority of genes are expressed at very low levels or not at all, while only a small percentage are highly active [@robinson2010].

### MDS by Experiment Batch {#MDS-by-Experiment-Batch}

```{r, fig.cap="Figure 3A/B: Multi-dimensional scaling (MDS) plot of pseudobulk samples. Distances correspond to leading log2-fold-changes between samples. (A) reveals a dominant batch effect between lab experiments, while (B) shows the subtle separation of GvHD and Control states."}
# Define factors for coloring
batch_colors <- as.numeric(factor(qc_summary$Batch))
group_colors <- as.numeric(factor(qc_summary$Group))

# Set up a 1x2 plotting area to show them side-by-side
par(mfrow=c(1,2))

# A. Experimental Clustering (The "Batch Effect")
plotMDS(dge, col = batch_colors, pch = 16, 
        main = "Figure 3A: Experimental Clustering",
        sub = "Colored by Batch")
legend("topleft", legend = levels(factor(qc_summary$Batch)), 
       col = 1:length(unique(qc_summary$Batch)), pch = 16, cex=0.7)

# B. Biological Clustering (The "Condition")
plotMDS(dge, col = group_colors, pch = 16, 
        main = "Figure 3B: Biological Clustering",
        sub = "Colored by Group")
legend("topleft", legend = levels(factor(qc_summary$Group)), 
       col = 1:length(unique(qc_summary$Group)), pch = 16, cex=0.7)

par(mfrow=c(1,1)) # Reset plotting area
```
To explore the underlying structure of the data and identify potential sources of variation, I performed Multi-Dimensional Scaling (MDS). By generating side-by-side plots colored by technical batch and biological condition, I evaluated the "Signal-to-Noise" ratio of the experiment. The technical clustering plot reveals that a significant portion of the variance is driven by the experiment batch, while the biological clustering plot shows how samples separate by GvHD status. This visualization is critical for justifying a "blocked" statistical model that can mathematically neutralize these technical effects during differential expression testing [@leek2010].

```{r, results='hide'}
 # 1. Extract the MDS object without plotting
mds_obj <- plotMDS(dge, plot = FALSE)

# 2. Create a diagnostic dataframe
mds_diagnostics <- data.frame(
  Sample_ID = colnames(dge),
  Dim1 = mds_obj$x,
  Dim2 = mds_obj$y,
  Group = qc_summary$Group,
  Batch = qc_summary$Batch,
  Lib_Size = colSums(dge$counts)
)

# 3. Calculate "Distance from Center" to help flag outliers statistically
# (Optional: helpful for your written interpretation)
group_means <- mds_diagnostics %>%
  group_by(Group) %>%
  summarise(mean_Dim1 = mean(Dim1), mean_Dim2 = mean(Dim2))

mds_diagnostics <- mds_diagnostics %>%
  left_join(group_means, by = "Group") %>%
  mutate(Dist_From_Group_Mean = sqrt((Dim1 - mean_Dim1)^2 + (Dim2 - mean_Dim2)^2)) %>%
  dplyr::select(-mean_Dim1, -mean_Dim2)

# 4. Print the Table
knitr::kable(mds_diagnostics, 
             digits = 3, 
             caption = "Table 2: MDS Diagnostics for Outlier Detection")
```
An analysis of MDS diagnostic coordinates confirms that the primary source of variance is technical, with a 11-log-unit shift on Dimension 1 between Experiment 2 and Experiment 3. While sample GSM7465919 shows a high distance from the group mean ($D = 5.254$), it was retained for three reasons:

1. Its library size (33.4M) is sufficient for analysis.The variance appears on Dimension 2, representing biological heterogeneity rather than technical failure.
2. With a total N=9, removing samples would severely compromise the power of the Quasi-Likelihood F-test.
3. Instead of removal, the batch effect was managed using a multi-factor design matrix ($\sim Batch + Group$) [@leek2010]. 



## Differential Gene Expression
```{r}
# 1. Align Metadata with DGEList 
# This ensures that your Batch and Group factors are in the exact same order as your counts
current_samples <- colnames(dge)
metadata_final <- metadata_clean[match(current_samples, metadata_clean$geo_accession), ]

#  2. Define Factors and Design Matrix 
batch_factor <- factor(metadata_final$Batch)
group_factor <- factor(metadata_final$Group)

# The Blocked Design: accounting for Batch while testing for Group
design <- model.matrix(~ batch_factor + group_factor)
rownames(design) <- current_samples

# 3. Run edgeR Quasi-Likelihood Pipeline 
# Estimate dispersion (how much the genes vary)
dge <- estimateDisp(dge, design)

# Fit the GLM
fit <- glmQLFit(dge, design)

# Run the test for the last column (the Group effect)
# This creates the 'qlf' object you need for the plots
qlf <- glmQLFTest(fit, coef = ncol(design))

# 4. Prepare results table 
top_results <- topTags(qlf, n = 10)$table
knitr::kable(top_results, caption = "Table 3: Top 10 Differentially Expressed Genes (GvHD vs. Control). Statistical output for the top-ranked genes, including $Log_2$ Fold Change, Likelihood Ratio/F-statistics, raw p-values, and False Discovery Rate (FDR). PNMA8C is identified as the primary significant biomarker ($FDR = 0.036$).")
```
The final analysis phase utilizes the Quasi-Likelihood (QL) F-test framework in edgeR to identify significant biomarkers. After aligning the clinical metadata with the expression matrix, I constructed a Blocked Design Matrix ($~Batch + Group$) to account for technical variance while specifically testing for the GvHD effect. The pipeline involves estimating genewise dispersions and fitting a generalized linear model (GLM) to the data. This rigorous approach identified PNMA8C as the primary significant biomarker ($FDR = 0.036$), proving its robustness despite the high technical noise present in the clinical dataset [@benjamini1995].


**How many genes were significantly differentially expressed?**

```{r}
# 1. Extract ALL genes from test result
all_genes <- topTags(qlf, n = nrow(qlf))$table

# 2. Filter for those that meet FDR < 0.05
sig_genes <- all_genes[all_genes$FDR < 0.05, ]

# 3. Get the count
num_sig <- nrow(sig_genes)
print(paste("Number of significant genes:", num_sig))
```


**What thresholds did you use and why?**

* Calculation: P-values were calculated using the Quasi-Likelihood (QL) F-test in edgeR, which is specifically designed to handle the uncertainty in small sample sizes [@edger2025].
* Thresholds: I used an FDR < 0.05 and |logFC| > 1. The logFC threshold ensures biological relevance (at least a 2-fold change), while the FDR threshold controls for the expected proportion of false positives [@benjamini1995].
* Results: Only 1 gene (PNMA8C) met both thresholds.

**Which method did you use? And Why?**

* Method: The Benjamini-Hochberg (BH) method was used to calculate the False Discovery Rate (FDR) [@benjamini1995].
* Why: It is the standard approach for RNA-seq because it balances the need to reduce false discoveries without being so overly strict (like Bonferroni) that it wipes out real biological signals.

**How many genes passed correction?**

* Results: Only one gene (PNMA8C) passed the correction with an $FDR = 0.036$ [@benjamini1995].
* Reflection: While finding only one gene might seem low, it represents the high degree of technical noise in the dataset [@leek2010]. It is a scientifically honest result that prioritizes integrity over volume.


### Volcano Plot

```{r, fig.cap="Figure 4: Volcano Plot of Differential Gene Expression. A scatter plot showing the relationship between $Log_2$ Fold Change (x-axis) and statistical significance ($-log_{10}$ p-value, y-axis). The red point denotes the significant differentially expressed gene (FDR < 0.05), PNMA8C. Grey points represent genes that did not meet the statistical threshold after Benjamini-Hochberg correction."}
# Create the 'Significant' status based on standard thresholds
de_results <- topTags(qlf, n = Inf)$table %>%
  tibble::rownames_to_column("SYMBOL") %>%
  mutate(Threshold = ifelse(FDR < 0.05 & abs(logFC) > 1, "Significant", "Non-Significant"))

# Plot
ggplot(de_results, aes(x = logFC, y = -log10(PValue), color = Threshold)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = c("grey", "red")) +
  theme_minimal() +
  geom_text(data = subset(de_results, FDR < 0.1), aes(label = SYMBOL), vjust = 1, hjust = 1) +
  labs(title = "Volcano Plot of GvHD vs Control",
       subtitle = "Red indicates FDR < 0.05 and |logFC| > 1",
       x = "Log2 Fold Change", y = "-Log10 P-Value")
```

The Lone Survivor (PNMA8C): The plot vividly illustrates why only one gene was labeled "Significant." While thousands of genes (grey dots) show high fold changes, only PNMA8C (the red dot) moved high enough on the Y-axis (significance) to cross the FDR < 0.05 threshold.

Significance vs. Magnitude: Notice the many grey dots on the far left and far right (some with a log2 fold change greater than 10). Despite their massive effect size, their low position on the Y-axis indicates they have high p-values. In a small, high-variance dataset like this, a gene can show a huge change in expression that is still deemed "non-significant" because the change isn't consistent across all replicates  [@slowikowski2024].

### Heat Map
```{r, fig.cap="Figure 5: Hierarchical Clustering and Heatmap of Top Differentially Expressed Genes. Heatmap showing Z-score normalized expression of the top 10 genes (ranked by p-value). Hierarchical clustering on the x-axis demonstrates that despite high technical noise, samples successfully cluster into GvHD and Control groups, revealing a distinct biological signature."}
library(pheatmap)

# 1. Extract logCPM values for the top 10 genes
top_gene_ids <- rownames(top_results)
logCPM_matrix <- cpm(dge, log = TRUE)
plot_data <- logCPM_matrix[top_gene_ids, ]

# 2. Create annotation bars for the top of the heatmap
annotation_col <- data.frame(
  Group = dge$samples$group,
  Batch = batch_factor
)
rownames(annotation_col) <- colnames(dge)

# 3. Plot
pheatmap(plot_data, 
         scale = "row", 
         annotation_col = annotation_col,
         main = "Expression Signature of Top 10 GvHD Genes",
         color = colorRampPalette(c("blue", "white", "red"))(100),
         cellwidth = 30, cellheight = 20)
```
The hierarchical clustering and heatmap of the top 10 differentially expressed genes provide a clear visualization of the transcriptomic signature separating the experimental groups. Despite the significant technical batch effects (Experiment_1, 2, and 3) previously identified in the MDS analysis, the samples successfully cluster into two primary branches corresponding to the GvHD (pink) and Control (salmon) conditions [@leek2010]. This indicates that the Blocked GLM design successfully isolated the biological signal from the technical noise. Within the heatmap, PNMA8C, TRAJ40, and COL4A6 exhibit strong up-regulation in the GvHD group, while genes like CXCL14 and FCGBP show a distinct down-regulation pattern compared to controls [@mccarthy2012]. The use of Z-score normalization (row scaling) highlights that these ten genes possess a consistent expression profile across replicates, reinforcing their potential as robust biomarkers for gastrointestinal GvHD progression in this cohort [@kolde2025]. 


# Interpret, and document 


## Why is the dataset of interest to you?

Acute Graft-vs-Host Disease (aGvHD) remains a leading cause of non-relapse mortality following bone marrow transplants. This dataset is particularly compelling because it captures the site-specific immune response within the colon at a high resolution. By converting the single-cell data into pseudobulk profiles, we can analyze the "aggregate" immune environment, which is often more reflective of the clinical pathology seen by doctors than looking at individual cells in isolation [@jarosch2023, @pseudobulk2020].

## What are the control and test conditions of the dataset?

Control: Colon biopsies from patients who underwent hematopoietic cell transplantation (HCT) but did not develop aGvHD (Grade 0).

Test: Colon biopsies from patients who developed Grade 2-4 acute GvHD. These samples represent the active inflammatory and destructive phase of the disease.

## How many samples in each of the conditions of your dataset?
The dataset was filtered to include only the gene expression (GEX) pools, excluding TCR-sequencing libraries to ensure a clean transcriptomic analysis.

Control: 3 samples

GvHD: 6 samples (across various levels of severity)
```{r}
table(metadata_clean$Group)
```


## Were there expression values that were not unique for specific genes? How did you handle these? 
Yes. It is common for multiple Ensembl IDs (different isoforms or genomic locations) to map to a single HUGO gene symbol. To handle this, I used a "Max-Expression" strategy.

By taking the maximum value across these redundant rows, I ensured that the most robust signal for that gene was preserved without artificially inflating the total count sums through addition. 

See: [Map to HUGO symbols](#Map-to-HUGO-symbols) 

```{r}
# Total rows in raw data
raw_row_count <- nrow(raw_counts)

# Mapping using AnnotationDbi
mapping <- select(org.Hs.eg.db, 
                  keys = rownames(raw_counts), 
                  column = "SYMBOL", 
                  keytype = "ENSEMBL")

# Calculating the loss
unmapped_count <- sum(is.na(mapping$SYMBOL))
final_unique_count <- nrow(clean_counts)
duplicates_handled <- (raw_row_count - unmapped_count) - final_unique_count 

print(paste("Number of duplicates:", duplicates_handled))

```

## Were there expression values that could not be mapped to current HUGO symbols? 

Yes. The raw 10x Genomics output contains many Ensembl IDs that correspond to non-coding RNAs, pseudogenes, or outdated identifiers that do not exist in the current org.Hs.eg.db database [@carlson2024]. These unmapped rows were discarded during the inner_join process to ensure the final matrix only contained high-confidence, standardized HUGO symbols suitable for pathway analysis. 

See: [Map to HUGO symbols](#Map-to-HUGO-symbols)  

```{r}

print(paste("Raw Ensembl IDs:", raw_row_count))
print(paste("Unmapped IDs removed:", unmapped_count))
print(paste("Percentage of data removed as 'noise':", 
            round((unmapped_count / raw_row_count) * 100, 2), "%"))
```

## Were there any outliers in your dataset? How were they handled in the originating paper? How many outliers were removed? 

During the MDS (Multi-Dimensional Scaling) diagnostic phase, I identified GSM7465919 as a significant outlier, with a distance of $D = 5.254$ from the main cluster.
* The technical observation: This sample was a GvHD biopsy that sat far away from its group members on Dimension 1, which was heavily influenced by the technical batch effect [@leek2010].

Ultimately, I chose to *retain all samples*, including GSM7465919, for several strategic reasons:
* Statistical Power: With only 9 biological replicates (3 Control, 6 GvHD), removing even one sample would significantly reduce the power of the model to detect changes.
* Accounting for Noise: Rather than deleting the outlier, I utilized a *Blocked Linear Model* ($~Batch + Group$). This allowed the math to "ignore" the batch-driven variance of the outlier while still using its biological data [@leek2010, [@mccarthy2012]].
* Scientific Integrity: GvHD is a heterogeneous disease. I concluded that GSM7465919 likely represented the true clinical variance found in human patients rather than a lab error.
* The Result: While this decision led to a very conservative result (only one significant gene, $PNMA8C$), it ensured that the result was highly robust and not the product of "p-hacking" by removing inconvenient data points.

See: [MDS by Experiment Batch](#MDS-by-Experiment-Batch)

The originating paper didn't remove any samples. They handled outliers primarily during the quantification of cell densities within the ChipCytometry dataset. In the linear regression analysis of Treg and CD8 T cell densities, the researchers explicitly noted the removal of only 8 outliers that exceeded $1.5 \times$ the interquartile range (IQR) [@leek2010].  

## How did you handle replicates? 

I utilized Pseudobulk Aggregation [@pseudobulk2020] . Since scRNA-seq data often suffers from "dropout" (zero counts) and "p-value inflation" (treating 1,000 cells from one person as 1,000 independent samples), summing the counts into a single profile per patient pool is the current best practice. This creates "true" biological replicates that allow for the use of robust bulk RNA-seq tools like edgeR [@edger2025]. 

See: [Prepare Dataset](#Prepare-Dataset) 

## What is the final coverage of your dataset? 

The final cleaned and normalized dataset contains 24,492 unique HUGO gene symbols across 9 biological samples. 

```{r}
nrow(clean_counts)
```

# References
